### 2023-07-07 records/manuscript_2023.tex
polish figures for manuscript & write first version of results

\section*{Introduction}
%
XXX
%
Behavior of intelligent life is context-dependent and requires a constant adaptation to the ever-changing life. This must also be reflected in the underlying neural networks and the inputs that drive those. For instance, when you walk down your your basement stairs and the lighting is well, your brain might rely entirely on the feedforward sensory input your senses receive to take decisions and adapt your body in the environment/world (Fig 1a). However, when you walk down the same basement stairs you walked million of times before but suddenly the light bulb is broken and the visual input is absent, you don't freeze forever but walk the stairs almost as confidently as before. Your brain now relies on the prediction it had formed over the last years. (Fig. 1b). The former example resonates with the feedforward-centric view that has dominated the neuroscience literature for decades. However, in recent years, the importance of internally generated, feedback inputs has increased significantly. Not only because it was shown that top-down projects xxx outsize feedforward connections (ref) but also because these connection modulate (ref) or even drive (ref) neuron dynamics. These feedback projects have been hypothesized to carry expectations, predictions ... and may help us to navigate (in the physical and metaphysical sense) the world even in the absence of sensory input.  But how do our brains decide whether it follows more the sensory feedforward input or the feedback predictions?

For instance, if you hike down a mountain for the first time under very foggy conditions, you brain may receive unreliable sensory input but at the same time can only have a very rough prediction about what to expect. In this case, your brain might be better off combining both the sensory input and the prediction thereof (Fig 1c). A common hypothesis is that the brain weights those signals according to their reliabilties, that is, a function of the variances of those signals (ref). These ideas come from Bayesian integration in which it is believed that xxx (ref). A prominent example of this is multisensory integration in which xxx (ref). Similarly, the brain must have mechanisms to estimate the variances of both the sensory input and the predictions thereof that allow it to weight those signals accordingly and context-dependent. But how this is implemented in the brain is still largely unknown.

Here, we hypothesis that xxx. 
We show that xxx

@@

\subsection*{nPE and pPE neurons as the basis for computing mean and variance of sensory stimuli}
%
We hypothesise that the distinct response patterns of negative and positive prediction-error (nPE/pPE) neurons represent the backbone for estimating the mean and the variance of sensory stimuli. nPE neurons only increase their activity relative to a baseline when the sensory input is weaker than predicted, while pPE neurons only increase their activity relative to a baseline when the sensory input is stronger than predicted. Moreover, both nPE and pPE neurons remain at their baseline activity when the sensory input is fully predicted (XXX). Assuming that the prediction equals the mean of the sensory stimulus, the PE neurons, hence, encode the deviation from the mean. Thus, the squared sum of nPE and pPE activity represents the variance of the feedforward input.

@@

\subsection*{Weighting external and internal signals requires two sets of PE neurons}
%
Following the ideas of Bayesian multisensory integration (XXX), the weighting of sensory stimuli and predictions thereof would require knowledge of their variances. As we have shown in the previous section, the variance of the sensory stimulus can be estimated by means of PE neurons. We hypothesise that the same principles apply to estimating the variance of the prediction. Hence, we augment the network with a \textit{higher} PE circuit that receives output synapses from the M neuron of the \textit{lower} PE circuit (Fig. 1xxx). Both subnetworks are modeled the same, except that the M neuron in the higher PE circuit evolves more slowly than the M neuron in the lower PE circuit.

@@

For some psychiatric disorders, it has been shown that the weighting of sensory inputs and predictions thereof is impaired (XXX), leading to an overweighting of one of these signals. Moreover, it has been hypothesised that factors like stress or cognitive load may also influence the processing of feedforward and feedback inputs (XXX).  Naturally, we were wondering which factors might influence the estimation of the variances, and, consequently, the weighting of different input streams. In our network, the M neuron evolves faster in the lower PE circuit than in the higher PE circuit. When we xxxx, the network shows a bias for predictions, while when we xxx, the network shows a bias for the sensory stimuli (Fig. SXX). This indicates that a non-pathological weighting requires xxx. Another factor that may contribute to a distorted weighting is the baseline activity of PE neurons that was set to zero in our model, in line with very low spontaneous firing rates of excitatory neurons in rodent primary sensory areas (XXX). Increasing this baseline activity for the nPE neuron (Fig. SXXX), pPE neuron (Fig. SXXX) or both pushes the network to weight sensory stimuli and predictions more equally. We speculate that an increase of baseline activity may be a natural result of an increased cognitive load or stress.

XXX Summary XXX
% I haven't talked about the different activation function but I don't see where this fits naturally ...

@@

\subsection*{XXX IN influence, neuromodulators ... mechanisms XXX}
%
The brain's flexibility and adaptability are not least due to the fact that a plethora of neuromodulators influence the activity of neurons in a variety of ways (XXX). A prominent target of neuromodulatory inputs are inhibitory neurons (Cardin 2019, XXX). Moreover, distinct interneuron types are differently (in-)activated by those neuromodulators. For instance, it has been shown that xxx. We were therefore wondering if and how the weighting of sensory inputs and predictions thereof may be biased when neuromodulators activate distinct interneuron types.

To this end, we modeled the presence of neuromodulators by injecting an additional, excitatory input into one or two interneuron types. We reasoned that the network effect of a neuromodulator not only depends on the interneuron type it targets but also on the inputs this neuron receives and the connections it makes with other neurons in the network. We, therefore, tested three different mean-field networks that differ with respect to the distribution of sensory inputs and predictions onto the interneurons and the underlying connectivity. The commonality across those networks is that all of them exhibit an E/I balance of excitatory and inhibitory pathways onto the PE neurons (XXX). Across the different mean-field networks tested, activating a SOM or VIP neuron individually forces the networks to weigh both inputs more equally. As a consequence, in a sensory-driven input regime, predictions are overrated. Similarly, in a prediction-driven regime, sensory inputs are overrated. Interestingly, when both interneuron types are activated to the same degree, this effect disappears (Fig. 5axx, left). In contrast, activating PV neurons largely biases the network output towards predictions. This is even more pronounced when PV and SOM, or PV and VIP neurons are activated simultaneously (Fig. 5a, middle and right).

In the previous simulations, we assumed that a neuromodulator acts globally, that is, on the interneurons in both the lower and the higher PE circuit. While this is in line with experimental data showing that xxx (XXX), we note that neuromodulators may also act more locally. We hence tested how the weighting of sensory stimuli and predictions may change when a neuromodulator activates specific interneurons in the lower or higher PE circuit only. The effect of activating an interneuron type in the lower PE circuit on the sensory weight is mostly the opposite of activating the same interneuron in the higher PE circuit (Fig. SXXX). For instance, the sensory inputs are overrated when VIP neurons are activated in the higher PE circuit, while the prediction is overrated when VIP neurons in the lower PE circuit are activated. When VIP and SOM neurons are stimulated equally, the sensory weight remains unchanged, independently of which PE circuit is targeted by the neuromodulator.

What are the mechanisms that give rise to these effects? And how do the combined local changes give rise to the global one observed in our network simulations?

% Mention that for V neuron (higher) the darker line is actually not horizontal (unaffected)!
% Mention that in BL illustration you need a factor to increase visibility ... (in caption)

\subsection*{XXX contraction bias}
% 
The weighting of sensory inputs and predictions thereof manifests in all-day behavior, in the form of a phenomenon called \textit{contraction bias}. The contraction bias describes the tendency to overestimate sensory stimuli that are at the lower end of a stimulus distribution and to underestimate stimuli that are at the upper end of the same distribution (Fig. 5aXX). This \textit{bias towards the mean} has been reported in different species and modalities (XXX).

@@

>>>>

\section*{Introduction}
%
To survive in an ever-changing environment, animals must flexibly adapt their behavior based on previously encoded and novel information. This context-dependency must also be reflected in the information processing of neural networks underlying intelligent behavior. For instance, when you walk down some staircase in your fully lit basement, your brain might entirely rely on the feedforward (bottom-up) input your senses receive (Fig 1axxx). In contrast, when you walk down the same stairs in complete darkness, your brain might rely entirely on feedback (top-down) signals generated from a staircase model it has formed over previous experiences (Fig. 1bxx).

The importance of these feedback inputs has been emphasized by observations showing that top-down projections outnumber feedforward connections (XXX) and that they modulate (XXX) or even entirely drive (XXX) neuron activity. But how do neural networks switch between a feedforward-dominated and a feedback-dominated processing mode? And how do neural networks combine both input streams wisely? For instance, if you hike down a mountain for the first time in very foggy conditions, your brain receives unreliable visual information. In addition, it can only draw on a shaky prediction about what to expect (Fig. 1cxxx).

A common hypothesis is that the brain weights different inputs according to their reliabilities which are a function of the variances of those inputs (XXX). A prominent example of this is Bayesian multisensory integration in which the information from multiple modalities is optimally combined (XXX). According to this theory, the representation of an object perceived through various senses is a linear combination of the single-modality estimates, weighted by their reliabilities. This theory has been backed up by a number of observations showing that xxx (XXX). The same concepts can be employed for the weighting of sensory feedforward inputs and predictions thereof (XXX). A central point in this weighting is the estimation of the sensory and prediction variances that underlie the reliability of each input. However, how this is implemented in the brain is still largely unknown.

We hypothesised that prediction error (PE) neurons can provide the basis for the neural computation of variances. PEs are an integral part of the theory of predictive processing that states that the brain constantly compares incoming sensory information with predictions thereof. When those predictions are wrong, the resulting PEs allow the network to update the model of the world, thereby ensuring that the predictions are more accurate (XXX). Experimental evidence suggests that these PEs may be represented in the activity of distinct groups of neurons, termed PE neurons (XXX). In brain areas with excitatory neurons that exhibit low spontaneous firing rates, these neurons may come in two types (XXX). Negative PE (nPE) neurons mainly increase their activity when the prediction is \textit{stronger} than the sensory input, while positive PE (pPE) neurons mainly increase their activity when the prediction is \textit{weaker} than the sensory input. Indeed, it has been shown that excitatory neurons in layer 2/3 of rodent primary sensory areas can encode negative or positive PEs (XXX).

% also, mention that PE neurons are formed by inhibition, so that you can mention your work and introduce different types of interneurons!? (maybe combine with below section, like as it has been shown that PE neurons depend on inhibition and E/I balance, we reasoned that neuromodulators acting on interneurons can hence bias the weighting and hence investigated  how this is infleunced .... something like that?)

Here, we show that

- unique response profiles of nPE and pPE can be the backbone for computing mean and variance of sensory stimuli
- Estimating both the variance of the sensory inputs and the prediction thereof requires a hierarchy of PE neurons
- In line with the theory, predictons are weighted more strongly than sensory information when environement is stable and therefor ethe prediction is reliable while the sensory stimuli are noisy
- moreover, even when sensory information are crystal clear but the environement, that is the stimuli shown, is unstable, leading to outweighing of sensory inputs, right at the beginning of  a trial, predictions kick in shortly
- neuromodulators acting globally on VIP neurons ..., acting on SOM neurons ... and acting on PV neurons ...
- we show that this is due to an increase/decrease of gain/baseline of PE neurons and can be predicted from effect of IN on both properties
- Finally, we show that the contraction bias is a manifestation of this weighting. We show that it depends on several factors, trail duration, depends on both the volatility of the environement and the stimulus uncertainty. A mixture of recent (last stimulus) if xxx and a more averaged lomger history of stimuli when xxx
- summary sentence

@@

\subsection*{nPE and pPE neurons as the basis for estimating mean and variance of sensory stimuli}
%
We hypothesise that the distinct response patterns of negative and positive prediction-error (nPE/pPE) neurons represent the backbone for estimating the mean and the variance of sensory stimuli. nPE neurons only increase their activity relative to a baseline when the sensory input is weaker than predicted, while pPE neurons only increase their activity relative to a baseline when the sensory input is stronger than predicted. Moreover, both nPE and pPE neurons remain at their baseline activity when the sensory input is fully predicted (XXX). Assuming that the prediction equals the mean of the sensory stimulus, the PE neurons, hence, encode the deviation from the mean. Thus, the squared sum of nPE and pPE activity represents the variance of the feedforward input.

@@

\subsection*{Estimating variances of sensory inputs and predictions requires a hierarchy of PE circuits}
%
Following the ideas of Bayesian multisensory integration (XXX), the weighting of sensory stimuli and predictions thereof would require knowledge of their variances. As we have shown in the previous section, the variance of the sensory stimulus can be estimated by means of PE neurons. We hypothesise that the same principles apply to estimating the variance of the prediction. Hence, we augment the network with a \textit{higher} PE circuit that receives output synapses from the M neuron of the \textit{lower} PE circuit (Fig. 1xxx). Both subnetworks are modeled the same, except that the M neuron in the higher PE circuit evolves more slowly than the M neuron in the lower PE circuit.

@@

For some psychiatric disorders, it has been shown that the weighting of sensory inputs and predictions thereof is impaired (XXX), leading to an overweighting of one of these signals. We thus wondered which network properties might bias the estimation of the variances, and, consequently, the weighting of different input streams. In our network, the M neuron evolves faster in the lower PE circuit than in the higher PE circuit. We identified the speed at which the M neurons are updated in the different PE circuits as a decisive network factor. To show this, we varied the weights from the PE neurons onto the M neurons in either PE circuit. When xxxx, the network shows a bias for predictions, while when xxx, the network shows a bias for the sensory stimuli (Fig. SXX). While the speed at which the activity of the M neurons evolve in relation to each other may underlie pathological weighting, the precise activation function of the V neurons is of less importance. When we replaced the squared activation function with a purely linear function, the V neurons do not encode the variance but the averaged absolute deviation of the sensory stimuli. However, the sensory weight is only slightly shifted to larger values for low trial/high stimulus variability (Fig. SXXX).

@@

\subsection*{Biasing the weighting of sensory inputs and predictions by neuromodulators acting on PE circuits}
%
The brain's flexibility and adaptability are not least due to the fact that a plethora of neuromodulators influence the activity of neurons in a variety of ways (XXX). A prominent target of neuromodulatory inputs are inhibitory neurons (Cardin 2019, XXX). Moreover, distinct interneuron types are differently (in-)activated by those neuromodulators. For instance, it has been shown that xxx. We therefore wondered if and how the weighting of sensory inputs and predictions thereof may be biased when neuromodulators activate distinct interneuron types.

To this end, we modeled the presence of neuromodulators by injecting an additional, excitatory input into one or two interneuron types. We reasoned that the network effect of a neuromodulator not only depends on the interneuron type it targets but also on the inputs this neuron receives and the connections it makes with other neurons in the network. We, therefore, tested three different mean-field networks that differ with respect to the distribution of sensory inputs and predictions onto the interneurons and the underlying connectivity. The commonality across those networks is that all of them exhibit an E/I balance of excitatory and inhibitory pathways onto the PE neurons (XXX). Across the different mean-field networks tested, activating a SOM or VIP neuron individually forces the networks to weigh both inputs more equally. As a consequence, in a sensory-driven input regime, predictions are overrated. Similarly, in a prediction-driven regime, sensory inputs are overrated. Interestingly, when both interneuron types are activated to the same degree, this effect disappears (Fig. 5axx, left). In contrast, stimulating PV neurons largely biases the network output towards predictions. This is even more pronounced when PV and SOM, or PV and VIP neurons are activated simultaneously (Fig. 5a, middle and right).

In the previous simulations, we assumed that a neuromodulator acts globally, that is, on the interneurons in both the lower and the higher PE circuit. While this is in line with experimental data showing that xxx (XXX), we note that neuromodulators may also act more locally. We hence tested how the weighting of sensory stimuli and predictions may change when a neuromodulator activates specific interneurons in the lower or higher PE circuit only. The effect of stimulating an interneuron type in the lower PE circuit on the sensory weight is mostly the opposite of activating the same interneuron in the higher PE circuit (Fig. SXXX). For instance, the sensory inputs are overrated when VIP neurons are activated in the higher PE circuit, while the prediction is overrated when VIP neurons in the lower PE circuit are stimulated. When VIP and SOM neurons are stimulated equally, the sensory weight remains unchanged, independently of which PE circuit is targeted by the neuromodulator.

What are the mechanisms that give rise to these effects? And how do the combined local changes give rise to the global one observed in our network simulations? The sensory weight is chosen to be a function of the V neurons of the lower and higher PE circuit. Hence, any changes to the sensory weight are a consequence of changes to the neurons encoding the variance (Fig. 5xx). In our network, the V neurons only receive excitatory output synapses from both nPE and pPE neurons. Hence, any changes in the sensory weights upon activation of interneurons must be due to changes in the PE neurons. To disentangle the effect of nPE and pPE neurons, we perturbed those neurons individually in both the lower or higher subnetwork by injecting either an inhibitory or excitatory additional input (Fig. 5cxxx). Stimulating either PE neuron in the lower subnetwork increases the activity of the lower V neuron strongly. Moreover, the higher V neuron is also slightly affected. At first, this is counterintuitive because the V neuron in the higher subnetwork does not receive direct synapses from the PE neurons in the lower subnetwork. However, the activity of the M neuron encoding the prediction increases with an excitatory input onto the pPE neuron and decreases when the same input targets the nPE neuron (the opposite is true for an inhibitory input). Because neurons in the higher PE circuit receive synapses from the M neuron of the lower subnetwork, the estimation of the variance in the higher subnetwork is also affected. In contrast, stimulating either PE neuron in the higher subnetwork increases the activity of the higher V neuron but leaves the M and V neurons in the lower subnetwork unaffected (Fig. 5cxxx).

The stimulation of the PE neurons may cause both an increase in the baseline activity and a change in the neuron's gain. To disentangle the effect of elevated baseline activity and gain change, we illustrate each contribution separately by means of a mathematically tractable toy model (see Methods for more details). The variance estimated in the lower subnetwork increases for both an increase in the baseline and the gain of the PE neurons in the lower subnetwork. In contrast, when the gain of those PE neurons decreases, so does the variance. As we have seen before, changes in the baseline and the gain of PE neurons in the higher PE circuit do not affect the lower-level variance estimation. The variance estimated in the higher subnetwork changes equally for baseline increase or gain change of the PE neurons in the same subnetwork. However, in line with the perturbation experiments above, the higher-level variance estimation is affected by changes to the baseline or gain of the lower-level PE neurons as a result of the biased prediction. The variance increases slightly with increasing baseline or increasing deviation from a gain equal to 1 (Fig 5dxxx). Moreover, the mean of the sensory stimuli is overpredicted when the baseline or the gain for the pPE neuron increases, or when the gain of the nPE neuron decreases. In contrast, the mean of the sensory stimuli is underpredicted when the baseline or the gain of the nPE neuron increases, or when the gain of the pPE neuron decreases. In summary, any change in interneurons that cause an additional inhibitory input to the PE neurons reduces the gain of those neurons. Changes in the inhibitory neurons that cause an additional excitatory input to the PE neurons increases the baseline and potentially the gain of those neurons.

This suggests that to understand the effect of neuromodulators on the sensory weight, we need to investigate the effect of interneuron activation on baseline and gain of PE neurons. To this end, we stimulated either PV, SOM, or VIP neurons independently for all three mean-field networks and measured the changes in baseline and gain of both PE neuron types (Fig. 5exx). 
In all three networks tested, activating PV neurons decreases both quantities, leading to a decrease in the variance estimated (Fig. 5exx \& Fig. 5Sxxx). Activating SOM or VIP neurons decreases the overall gain but increases the baseline activity of the PE neurons. Whether and how much the gain of the nPE or pPE neuron is reduced depends on the inputs onto SOM/VIP neuron and the connectivity to the other neurons in the network. Similarly, how much the baseline is elevated depends on the specifics of the mean-field network (Fig. 5exx \& Fig. 5Sxxx). Hence, whether stimulating SOM or VIP neurons decreases or increases the activity of the V neurons depend on the input statistics. For low-mean stimuli, the effect of the elevated baseline dominates, while for high-mean stimuli the effect of the gain changes dominates the combined effect.

XXX Summary ... effect of neuromodulators acting on INs on sensory weight best understood as combined effect on the baseline and gain of nPE and pPE neurons that in turn determine the activity of the V neuron. Those effects not ony depend on the target IN but also the network in which it is embedded, that is the connectivity and the inputs onto INs. Moreover, it depends on whether local or global. The effects of activation on lower is mainly the opposite of same in higher. However, the efefcts don't cancel each other in global modulation. Mostly because 1) V higher additionally changes due to changes in prediction but also 2) effects in higher are stronger !? and 3) 1/(1+(Vs + D)/(Vp + D)) ... that is it cannot cancel per se. Even though the underlying reasons for a change in the V neuron depends on the the network characteristics, there are some consistent changes in sensory weigt across MFN ... name

% in caption: you need to make sure that in e you used a perturbationof 1 and the mean and std of stimuli were ... the lines only illustrate that it depends on the input statistics. Of course, the markers would also change with other perturbaton strengths, would they though? Think through
% Mention that for V neuron (higher) the darker line is actually not horizontal (unaffected)!
% Mention that in BL illustration you need a factor to increase visibility ... (in caption)

\subsection*{The contraction bias as a result of the weighting of sensory inputs and predictions thereof}
% 
The weighting of sensory inputs and predictions thereof manifests in all-day behavior, in the form of a phenomenon called \textit{contraction bias}. The contraction bias describes the tendency to overestimate sensory stimuli that are at the lower end of a stimulus distribution and to underestimate stimuli that are at the upper end of the same distribution (Fig. 5aXX). This \textit{bias towards the mean} has been reported in different species and modalities (XXX).

@@

% At the end, or in between talk about changes in PE neurons directly that you have currently in Fig. 3 SX ... I copy the part of the last paragraph here as inspiration :
% For some psychiatric disorders, it has been shown that the weighting of sensory inputs and predictions thereof is impaired (XXX), leading to an overweighting of one of these signals. Moreover, it has been hypothesised that factors like stress or cognitive load may also influence the processing of feedforward and feedback inputs (XXX). ... Another factor that may contribute to a distorted weighting is the baseline activity of PE neurons that was set to zero in our model, in line with very low spontaneous firing rates of excitatory neurons in rodent primary sensory areas (XXX). Increasing this baseline activity for the nPE neuron (Fig. SXXX), pPE neuron (Fig. SXXX) or both pushes the network to weight sensory stimuli and predictions more equally. We speculate that an increase of baseline activity may be a natural result of an increased cognitive load or stress.

% mention Pakan paper
